---
title: "Assignment 2: Coding Basics"
author: "Sammy DiLoreto"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
sequence.q1 <- seq(1, 100, 4) 
  #creating a sequence from one to 100, increasing by 4 (named sequence.q1)
sequence.q1 #returning sequence as output

#2. 
mean.q1 <- mean(sequence.q1) #computing the mean of sequence.q1
median.q1 <- median(sequence.q1) #computing the median of sequence.q1
mean.q1 #returning mean as output
median.q1 #returning median as output

#3. 
mean.q1 > median.q1 #determining if mean.q1 is greater than median.q1

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5.
student.names <- c("Jenny", "Grace", "Delaney", "Natalie") 
  #vector containing student names
test.scores <- c(49, 72, 97, 81) 
  #vector containing student's test scores (out of 100)
passed <- c(FALSE, TRUE, TRUE, TRUE) 
  #vector containing whether or no the students passed (passing grade of 50)

#6.
class(student.names) #character vector of student names
class(test.scores) #numeric vector of test scores
class(passed) #logical vector of whether or not they passed

#7. and 8. 
students.scores_df <- data.frame("StudentNames"=student.names,
                                 "TestScores"=test.scores, "Passed"=passed)
students.scores_df
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is a list of vectors with the same length but difference mode (character, numeric, logical) whereas a matrix has the same length and same mode. A matrix would have to consist of all numeric or all character, etc.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

> Possible Function1:
passing.grade <- function(test.scores){
  output <- ifelse(test.scores>=50, TRUE, FALSE)
  print(output)}
passing.grade(test.scores)

> Possible Function2:
passing.grade2 <- function(test.scores){
  output <- if(test.scores>=50) {TRUE} 
  else {FALSE}
  print(output)}
passing.grade2(test.scores)

  
11. Apply your function to the vector with test scores that you created in number 5.

```{r}
passing.grade <- function(test.scores){
  output <- ifelse(test.scores>=50, TRUE, FALSE)
  print(output)}
passing.grade(test.scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "ifelse" worked but "if" and "else" does not because "if" cannot run all values in a vector, it can only run a single value. A "for" loop would be needed to run all values in the vector. 
