---
title: "Assignment 10: Data Scraping"
author: "Sammy DiLoreto"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, message = FALSE, warning = FALSE}
#1 Set up session
library(tidyverse)
library(lubridate)
library(here) 
library(rvest)
getwd()
here()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website, message = FALSE, warning = FALSE}
#2 set variables for components of url
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php'
the_pwsid <- '03-32-010'
the_year <- 2022
the_scrape_url <- paste0(the_base_url, '?pwsid=', the_pwsid, '&year=', the_year)
print(the_scrape_url)

the_website <- read_html(the_scrape_url)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data, message = FALSE, warning = FALSE}
#3 assign variables to the values we want to scrape
water.system.name <- the_website %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
  html_text()

PWSID <- the_website %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% 
  html_text()
  
ownership <- the_website %>% 
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% 
  html_text()

max.withdrawals.mgd <- the_website %>% 
  html_nodes('th~ td+ td') %>% 
  html_text()
  
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the average daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, message = FALSE, warning = FALSE}
#4 convert scraped data into dataframe
durham.2022.df <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 9, 4, 8, 12),
                              "Year" = the_year,
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>% 
  mutate(WaterSystemName = !!water.system.name,
         PWSID = !!PWSID,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))

#5 Create a line plot of the max daily withdrawals across the months for 2022
Max.Daily.2022 <- ggplot(durham.2022.df, 
       aes(x= factor(Month, levels = 1:12, labels = month.abb),
         y=Max_Withdrawals_mgd,
         group=1)) + 
  geom_line() +
  labs(x = "Month", 
       y="Max Daily Use (MGD)", 
       title = "Max Daily Withdrawals in Durham in 2022" )
Max.Daily.2022

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function, message = FALSE, warning = FALSE}
#6. Construct function
#Create our scraping function
scrape.it <- function(the_year, the_pwsid){
  
  #Retrieve the website contents 
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php',
                                  '?pwsid=', the_pwsid, '&year=', the_year))
  
  #Scrape the data items
  water.system.name <- the_website %>% 
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
    html_text()

  PWSID <- the_website %>% 
    html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% 
    html_text()
    
  ownership <- the_website %>% 
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% 
    html_text()
  
  max.withdrawals.mgd <- the_website %>% 
    html_nodes('th~ td+ td') %>% 
    html_text()
  
  #Convert to a dataframe
  water.supply.df <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 9, 4, 8, 12),
                              "Year" = the_year,
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>% 
  mutate(WaterSystemName = !!water.system.name,
         PWSID = !!PWSID,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))
  
  #Return the dataframe
  return(water.supply.df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, message = FALSE, warning = FALSE}
#7 
#extract max daily withdrawals for Durham for each month in 2015
durham.2015.df <- scrape.it(2015,'03-32-010')

#plot it
Max.Daily.2015 <- ggplot(durham.2015.df, 
       aes(x= factor(Month, levels = 1:12, labels = month.abb),
         y=Max_Withdrawals_mgd,
         group = 1)) + 
  geom_line() +
  labs(x = "Month", 
       y="Max Daily Use (MGD)", 
       title = "Max Daily Withdrawals in Durham in 2015")
Max.Daily.2015

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, message = FALSE, warning = FALSE}
#8 
#extract data for Asheville (PWSID = 01-11-010) in 2015
asheville.2015.df <- scrape.it(2015,'01-11-010')

#combine with Durham 2015 data
ash.durm.2015.df <- rbind(durham.2015.df, asheville.2015.df)

#plot as separate lines to compare
ggplot(ash.durm.2015.df, 
       aes(x= factor(Month, levels = 1:12, labels = month.abb),
           y= Max_Withdrawals_mgd,
         color= WaterSystemName,
         group= WaterSystemName)) + 
  geom_line() +
  labs(x = "Month", 
       y="Max Daily Use (MGD)", 
       title = "Max Daily Withdrawals in Durham vs Asheville in 2015", 
       color= "Water System Name")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, message = FALSE, warning = FALSE }
#9 
#Set the inputs to scrape years 2015 to 2020 for the site "0004-0001"
the_years = rep(2010:2021)
my_pwsid = '01-11-010'

#Use lapply to apply the scrape function
ash.2010.2021.dfs <- lapply(X = the_years,
                  FUN = scrape.it,
                  the_pwsid=my_pwsid)

#Conflate the returned dataframes into a single dataframe
ash.2010.2021.df <- bind_rows(ash.2010.2021.dfs)

#plot it
ggplot(ash.2010.2021.df,aes(x=Date,y=Max_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE)+
  labs(x = "Year", 
       y="Max Daily Use (MGD)", 
       title = "Max Daily Withdrawals in Asheville (2010-2021)")


```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>It appears that Asheville has been increasing their water usage since 2015. Between 2010 and 2015 they decreased their water usage slightly but have been increasing since then. 
